---
title: "Case Study 2"
author: "Y Kunar, J Lancon, J Lubich, W Holt"
date: "December 04, 2017"
toc: true
output:
  prettydoc::html_pretty:
    theme: tactile
    highlight: github
    fig_caption: yes
---

### Project Name: EDA - Attrition Data Analysis
### Client: DDSAnalytics
#### Project Description: Preliminary Exploratory Data Analysis - (Employee Retention)
##### GitHub Repository: https://github.com/jlubich/DDS-CASE-STUDY-2

Header 1
---------------
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;DDSAnalytics is an analytics company that specializes in talent management solutions for Fortune 1000 companies. Having access to valuable demographic information obtained from employee's personnel files and workplace environmental data (Employee Surveys), DDSAnalytics executive management is interested leveraging this wealth of data, to provide additional services to their client base and assist them in improving their talent management strategies.  
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;DDSAnalytics' management understands the key role an excellent talent management system play in the development of any company.  Untold amounts of money and resources are spent hiring, training, and developing the most valuable asset a company has; Their Employees. The path is fraught with missteps, from hiring the 'wrong fit' personnel, not identifying and nurturing young talent, loss of productivity due to inexperienced personnel and lower profits due to excessive expenses associated with recruiting new employees.  
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;**Purpose:** Through the use of Data Science, DDSAnalytics identified an opportunity to gain a competitive edge by helping their clients better manage their talent. Before management green lights the development of the data science department, they have tasked the data science(DS) team to conduct an analysis of existing employee data.  The executive leadership has identified predicting employee turnover as its first application for the data science initiative.
The DS team was presented an existing dataset, to conduct an exploritory data analysis (EDA), to determine factor that leads to employee attrition. The major objectives are as follows:  

* Identify and discuss the top factors that contribute to employee attrition.  
* Determine if any job role specific trends that may exist.  
* Identify any interesting trends and/or observations  
* Determine Next-Steps in EDS process.  

**Dataset**
  
* **Client Dataset:** The DS team was presente with an annimus client's employee dataset. Certain identifiyable information was removed from the dataset (employee names, company name, etc.).  The dataset contains 35 variables with 1470 observations. The variables captured both employee data (age, sex, tenure, position, wages, performance reviews, education,...) as well as workplace envionmental data (Job involvement, job satisfaction, work/life balance, travel%,...), which provides a more complete picture of the workers and their environment.  
**Dataset:**  CaseStudy2Data.csv


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```
```{r DirectorySet, eval=FALSE}
## Initialize Directories and Loading Libraries
library(plyr)
library(dplyr)
library(binr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(tidyr)
library(ggplot2)
library(mapproj)
library(grid)
library(gridExtra)


setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;Data Visualization/Exploration:  The first step in the analysis process is to load the dataset and review its contents.  The best methods for this are visually looking at the data.  
**Fig 1.1 Variables - Classifications/Observations:** contains the variable name, variable class, and first 10 observations.  
**Table 1.2 Summary Statistics:** contains statistical summaries of each catagort (mean, median, min/max, etc.).  This give the team a high-level understanding of what is included in the dataset.

: **Fig 1.1** (Variables - Classifications/Observations)  
```{r DataLoading_n_Exploring}
##############################################################################
## 1 - Explore the available data and clean as necessary
##############################################################################


## Load up the Talent Data
TalentData <- read.csv("./DATA/casestudy2-data.csv")

## Review basic summary of data contents
str(TalentData)
```

: **Table 1.2** (Summary Statistics)
```{r SummaryData}
SumTalentData <- summary(TalentData)
print(SumTalentData)
```
A quick review of the aummary data tables above shows some basic statistics about the company and it's employees.  
* Attrition: 237 of 1470 = 16.1%  
* Gender: M 882 : F 588  60%:40% split  
* Age Range: 18 to 60 with median of 36 yrs old  
* Years of Tenure: 0 to 40 yrs with median of 5.0 yrs  
* Professional Company with Research & Development and Sales Roles dominating the ranks  

With just this initial data, it is possible to begin to draw a picture of the company and it's employees.

#### **Data Cleaning & Manipulation**
  
**Data Cleaning:** Reviewing the dataset shows the data is very clean and contains very few missing observations. 

**Data Manipluatoin/Munging:** 
human readable
```{r Labeling}
## Fix Garbled Column Name
names(TalentData)[1] <- "Age"

## Recode Attrition flag for counting
TalentData$AttritionCount <- ifelse(TalentData$Attrition=="Yes",1,0)

## Map numerical flags to Categorical Labels
TalentData$Education.Label <- plyr::mapvalues(
  TalentData$Education, 
  from = c(1,2,3,4,5), 
  to = c("Below College", "College", "Bachelor", "Master", "Doctor")
)

TalentData$EnvironmentSatisfaction.Label <- plyr::mapvalues(
  TalentData$EnvironmentSatisfaction, 
  from = c(1,2,3,4), 
  to = c("Low", "Medium", "High", "Very High")
)

TalentData$JobInvolvement.Label <- plyr::mapvalues(
  TalentData$JobInvolvement, 
  from = c(1,2,3,4), 
  to = c("Low", "Medium", "High", "Very High")
)

TalentData$JobSatisfaction.Label <- plyr::mapvalues(
  TalentData$JobSatisfaction, 
  from = c(1,2,3,4), 
  to = c("Low", "Medium", "High", "Very High")
)

TalentData$PerformanceRating.Label <- plyr::mapvalues(
  TalentData$PerformanceRating, 
  from = c(1,2,3,4), 
  to = c("Low", "Good", "Excellent", "Outstanding")
)

TalentData$RelationshipSatisfaction.Label <- plyr::mapvalues(
  TalentData$RelationshipSatisfaction, 
  from = c(1,2,3,4), 
  to = c("Low", "Medium", "High", "Very High")
)

TalentData$WorkLifeBalance.Label <- plyr::mapvalues(
  TalentData$WorkLifeBalance, 
  from = c(1,2,3,4), 
  to = c("Bad", "Good", "Better", "Best")
)


# set up boundaries for intervals/bins
breaks <- c(0,25,35,45,55,65,Inf)
# specify interval/bin labels
labels <- c("18-24", "25-34", "35-44", "45-54", "55-64", "65+")
# bucketing data points into bins
TalentData$Age.Bin <- cut(TalentData$Age, breaks, include.lowest = T, right=FALSE, labels=labels)

# set up boundaries for intervals/bins
breaks <- c(0,5,10,15,20,25,30,35,40,Inf)
# specify interval/bin labels
labels <- c("0-4", "5-9", "10-14", "15-19", "20-24", "25-29", "30-34", "35-39", "40+")
# bucketing data points into bins
TalentData$DistanceFromHome.Bin <- cut(TalentData$DistanceFromHome, breaks, include.lowest = T, right=FALSE, labels=labels)

# set up boundaries for intervals/bins
breaks <- c(0,3,6,11,21,Inf)
# specify interval/bin labels
labels <- c("0-1", "2-5", "6-10", "11-20", "20+")
# bucketing data points into bins
TalentData$YearsAtCompany.Bin <- cut(TalentData$YearsAtCompany, breaks, include.lowest = T, right=FALSE, labels=labels)


# set up boundaries for intervals/bins
breaks <- c(0,3,6,11,21,Inf)
# specify interval/bin labels
labels <- c("0-1", "2-5", "6-10", "11-20", "20+")
# bucketing data points into bins
TalentData$TotalWorkingYears.Bin <- cut(TalentData$TotalWorkingYears, breaks, include.lowest = T, right=FALSE, labels=labels)

# set up boundaries for intervals/bins
breaks <- c(0,5000,10000,15000,20000,25000,Inf)
# specify interval/bin labels
labels <- c("0-5000", "5001-10000","10001-15000","15001-20000","20001-25000","25000+")
# bucketing data points into bins
TalentData$MonthlyRate.Bin <- cut(TalentData$MonthlyRate, breaks, include.lowest = T, right=FALSE, labels=labels)
ac <- table(TalentData$MonthlyRate.Bin)
```

### barplot(ac)









ac <- table(TalentData$EducationField)
barplot(ac)

?bins.split(TalentData$Age, 8, 0, 2000)
bins.quantiles(TalentData$Age, 8, 8)


?cut2split(TalentData, cut2(TalentData$Age, g=10))
 


##############################################################################
## 2 - Establish the baseline attrition rate
##############################################################################

EmployeeCount.Total = sum(TalentData$EmployeeCount)
AttritionCount.Total = sum(TalentData$AttritionCount)

AttritionRate.Percent = round(AttritionCount.Total / EmployeeCount.Total, 2) * 100 






##############################################################################
## 3 - Identify 3-4 Attrition Rate factors
##############################################################################


## We can create mini pivot tables that allow us to calculate a percentage for each category within a given variable.
## For example we can build a table that shows us the attrition rate of each BusinessTravel group as seen below:

library(tidyr)
library(dplyr)

### Read data into R
HR <- read.csv("./DATA/casestudy2-data.csv")

## Note: There is no need to convert yes and no to 1 and 0

### Turn desired variables into a table
HRtable <- table(HR$BusinessTravel, HR$Attrition)

#Obtain the rate of attrition for each group
round(prop.table(HRtable,1),2)

Attrition.List <- list()

for(i in names(HR)){
  HRtable2 <- table(HR[[i]], HR$Attrition)
  HRtable3 <- round(prop.table(HRtable2,1),2)
  #ColumnName <- as.character(names(HR)[i])
  #HRtable3$ColumnName <- ColumnName
  Attrition.List[[i]] <- HRtable3
}



## Here's the magic function which computes the Attrition Rates, and Counts
fun.AttritionRate <- function(input.table, group.by.column) {
  input.table %>%
    dplyr::group_by(CategoryVar = input.table[,group.by.column]) %>%
    dplyr::summarise(
      AttritionRate = sum(AttritionCount)/sum(EmployeeCount), 
      AttritionCount = sum(AttritionCount),
      EmployeeCount = sum(EmployeeCount)) %>%
    dplyr::arrange(CategoryVar)
}

## Choose the columns which make sense
ColumnsToAnalyze <- c(
  which( colnames(TalentData)=="NumberOfCompaniesWorked" ),
  which( colnames(TalentData)=="BusinessTravel" ),
  which( colnames(TalentData)=="Department" ),
  which( colnames(TalentData)=="EducationField" ),
  which( colnames(TalentData)=="Gender" ),
  which( colnames(TalentData)=="MaritalStatus" ),
  which( colnames(TalentData)=="JobRole" ),
  which( colnames(TalentData)=="OverTime" ),
  grep("*Label", names(TalentData)), 
  grep("*Bin", names(TalentData))
)


#TalentData <- as_tibble(TalentData)

## Initilize a list to hold all the attrition rate details
Attrition.List = list()

## Loop through all of the columns in TalentData and get the attrition rate details
for(i in ColumnsToAnalyze){
  #Get the name of the current column 
  SummaryColumn <- names(TalentData)[i]
  #Get the attrition rate details into a DF
  SummaryResults <- fun.AttritionRate(TalentData, i)
  #Concatenate the name of the current column to the details
  SummaryResults$ColumnName <- SummaryColumn
  #Store the details into a list with all other column details
  Attrition.List[[i]] <- SummaryResults
}  

## Create a single data frame out of all of the detail results
attrition_data = do.call(rbind, Attrition.List)


## Plot
ggplot(data = attrition_data, mapping = aes(x = CategoryVar, y = AttritionRate, width=AttritionCount/100)) + 
  geom_bar(stat="identity", position="identity") + 
  geom_hline(yintercept = .16) + 
  geom_text(aes(label=AttritionCount), vjust=1.6, color="white", size=3.5)+
  #theme(axis.text.x=element_text(angle=45,vjust=0)) +  
  facet_wrap(~ColumnName, scales = 'free_x', ncol = 6)

jobrole <- attrition_data[which(big_data$ColumnName == "JobRole"),]

## Plot
ggplot(data = jobrole, mapping = aes(x = CategoryVar, y = AttritionRate, width=AttritionCount/100)) + 
  geom_bar(stat="identity", position="identity") + 
  geom_hline(yintercept = .16) + 
  geom_text(aes(label=AttritionCount), vjust=1.6, color="white", size=3.5)+
  #theme(axis.text.x=element_text(angle=45,vjust=0)) +  
  facet_wrap(~ColumnName, scales = 'free_x', ncol = 6)



### Use the comments below if row% is not desired
### for cell %
## round(prop.table(DF$NAME),2)

### for row %
## round(prop.table(DF$NAME,1),2)

### for column %
## round(prop.table(DF$NAME,2),2)


## If we wanted to include something like hourly rate, even though we don't at this point, we can still use the above structure and then we need to find a way to group the data within the table. 

##############################################################################
## 4 - Identify 3-4 interesting Job Specific Trends
##############################################################################

##############################################################################
## 5 Conclusions - where focused effort can get you the biggest bang for the buck
##############################################################################

