---
title: "Case Study 2"
author: " W Holt, Y Kunar, J Lancon, J Lubich"
date: "December 04, 2017"
output:
  prettydoc::html_pretty:
    theme: tactile
    highlight: github
    fig_caption: yes
---

### Project Name: EDA - Attrition Data Analysis
### Client: DDSAnalytics
#### Project Description: Preliminary Exploratory Data Analysis - (Employee Retention)
##### GitHub Repository: https://github.com/jlubich/DDS-CASE-STUDY-2

Header 1
---------------
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;DDSAnalytics is an analytic company that specializes in talent management solutions for Fortune 1000 companies. Having access to valuable demographic information obtained from employee's personnel files and workplace environmental data (Employee Surveys), DDSAnalytics executive management is interested leveraging this wealth of data, to provide additional services to their client base and assist them in improving their talent management strategies.  
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;DDSAnalytics' management understands the key role an excellent talent management system play in the development of any company.  Untold amounts of money and resources are spent hiring, training, and developing the most valuable asset a company has; Their Employees. The path is fraught with missteps, from hiring the 'wrong fit' personnel, not identifying and nurturing young talent, loss of productivity due to inexperienced personnel and lower profits due to excessive expenses associated with recruiting new employees.  
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;**Purpose:** Through the use of Data Science, DDSAnalytics identified an opportunity to gain a competitive edge by helping their clients better manage their talent. Before management green lights the development of the data science department, they have tasked the data science(DS) team to conduct an analysis of existing employee data.  The executive leadership has identified predicting employee turnover as its first application for the data science initiative.
The DS team was presented an existing data-set, to conduct an exploratory data analysis (EDA), to determine factor that leads to employee attrition. The major objectives are as follows:  

* Identify and discuss the top factors that contribute to employee attrition.  
* Determine if any job role specific trends that may exist.  
* Identify any interesting trends and/or observations  
* Determine Next-Steps in EDA process.  

**Dataset**
  
* **Client Dataset:** The DS team was presented with an anonymous client's employee data-set. Certain identifiable information was removed from the data-set (employee names, company name, etc.).  The data-set contains 35 variables with 1470 observations. The variables captured both employee data (age, sex, tenure, position, wages, performance reviews, education,...) as well as workplace environmental data (Job involvement, job satisfaction, work/life balance, travel%,...), which provides a more complete picture of the workers and their environment.  
**Dataset:**  CaseStudy2Data.csv


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
```
```{r LibraryLoads, include=FALSE}
## Initialize Directories and Loading Libraries
library(plyr)
library(dplyr)
library(binr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(tidyr)
library(grid)
library(gridExtra)
```
```{r DirectorySett, eval=FALSE, include=FALSE}
## Initialize Directory
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```

***  

#### **Data Visualization/Exploration:**  

The first step in the analysis process is to load the data-set and review its contents.  The best methods for this are visually looking at the data.  

**Table 1.1 Variables - Classifications/Observations:** contains the variable name, variable class, and first 10 observations.  
: **Fig 1.1** (Variables - Classifications/Observations)  
```{r DataLoading_n_Exploring, include=TRUE}
##############################################################################
## 1 - Explore the available data and clean as necessary
##############################################################################


## Load up the Talent Data
TalentData <- read.csv("./DATA/casestudy2-data.csv")
HR <- TalentData
## Review basic summary of data contents
str(TalentData)
```
**Table 1.2 Summary Statistics:** contains statistical summaries of each category (mean, median, min/max, etc.).  This give the team a high-level understanding of what is included in the data-set.

: **Figure 1.2** (Summary Statistics)
```{r SummaryData}
SumTalentData <- summary(TalentData)
print(SumTalentData)
```
A quick review of the summary data tables above shows some basic statistics about the company and it's employees.  

* **Attrition:** 237 of 1470 = **16.1%**  
* **Gender:** M 882 : F 588  60%:40% split  
* **Age Range:** 18 to 60 with median of 36 yrs old  
* **Years of Tenure:** 0 to 40 yrs with median of 5.0 yrs  
* **Type of Organization:** Professional Company; Research & Development and Sales Roles dominating the ranks  

With just this initial data, it is possible to begin to draw a picture of the company and it's employees.

***

#### **Data Cleaning & Manipulation**
  
**Data Cleaning:** 
After reviewing the original data-set, the data appears very well organized and complete, with very few missing observations. 
```{r Labeling}
## Fix Garbled Column Name
names(TalentData)[1] <- "Age"
```
**Data Manipulation/Munging:** 
One new column has been derived to transform the Yes/No values of attrition into a count which we can use for further analysis and visualizations.
```{r }
## Recode Attrition flag for counting
TalentData$AttritionCount <- ifelse(TalentData$Attrition=="Yes",1,0)
```
**Data Catagorization:**  The data-set contains variables that are categorical in nature.  They appear to be excerpts from employee surveys and/or personnel performance reviews.  These variables are numerically rankings based (1,2,3,4) where 1 = Low, 2 = Med, etc..  A legend of what each categorical variable ranking means was included in the data-set.  
While this format is good data entry and manipulation, it is not 'human readable' and not very informative; therefore, the team has manipulated the data-set to display the 'human readable' variable format. We created a new variable for each categorical variable transformed and labeled it with suffix *'VariableName'*.Label.  

```{r}
## Map numerical flags to Categorical Labels
TalentData$Education.Label <- plyr::mapvalues(
  TalentData$Education, 
  from = c(1,2,3,4,5), 
  to = c("Below College", "College", "Bachelor", "Master", "Doctor")
)

TalentData$EnvironmentSatisfaction.Label <- plyr::mapvalues(
  TalentData$EnvironmentSatisfaction, 
  from = c(1,2,3,4), 
  to = c("Low", "Medium", "High", "Very High")
)

TalentData$JobInvolvement.Label <- plyr::mapvalues(
  TalentData$JobInvolvement, 
  from = c(1,2,3,4), 
  to = c("Low", "Medium", "High", "Very High")
)

TalentData$JobSatisfaction.Label <- plyr::mapvalues(
  TalentData$JobSatisfaction, 
  from = c(1,2,3,4), 
  to = c("Low", "Medium", "High", "Very High")
)

TalentData$PerformanceRating.Label <- plyr::mapvalues(
  TalentData$PerformanceRating, 
  from = c(1,2,3,4), 
  to = c("Low", "Good", "Excellent", "Outstanding")
)

TalentData$RelationshipSatisfaction.Label <- plyr::mapvalues(
  TalentData$RelationshipSatisfaction, 
  from = c(1,2,3,4), 
  to = c("Low", "Medium", "High", "Very High")
)

TalentData$WorkLifeBalance.Label <- plyr::mapvalues(
  TalentData$WorkLifeBalance, 
  from = c(1,2,3,4), 
  to = c("Bad", "Good", "Better", "Best")
)

```
**Data Binning:**  Several variables have either continuous numerical data (Income, travel distance,..) or numerous discrete values (age,..).  To better visualize, reduce clutter, and interpret the data interpret, the team created Bins/Data-Ranges for these variables.We created a new variable for each categorical variable transformed and labeled it with suffix *'VariableName'*.Bin

```{r }


# Age Bins#
# set up boundaries for intervals/bins
breaks <- c(0,25,35,45,55,65,Inf)
# specify interval/bin labels
labels <- c("18-24", "25-34", "35-44", "45-54", "55-64", "65+")
# bucketing data points into bins
TalentData$Age.Bin <- cut(TalentData$Age, breaks, include.lowest = T, right=FALSE, labels=labels)


# Distance-from-Home Bins#
# set up boundaries for intervals/bins
breaks <- c(0,5,10,15,20,25,30,35,40,Inf)
# specify interval/bin labels
labels <- c("0-4", "5-9", "10-14", "15-19", "20-24", "25-29", "30-34", "35-39", "40+")
# bucketing data points into bins
TalentData$DistanceFromHome.Bin <- cut(TalentData$DistanceFromHome, breaks, include.lowest = T, right=FALSE, labels=labels)


# Years at Company Bins#
# set up boundaries for intervals/bins
breaks <- c(0,3,6,11,21,Inf)
# specify interval/bin labels
labels <- c("0-1", "2-5", "6-10", "11-20", "20+")
# bucketing data points into bins
TalentData$YearsAtCompany.Bin <- cut(TalentData$YearsAtCompany, breaks, include.lowest = T, right=FALSE, labels=labels)


# Total-Working-Years Bins#
# set up boundaries for intervals/bins
breaks <- c(0,3,6,11,21,Inf)
# specify interval/bin labels
labels <- c("0-1", "2-5", "6-10", "11-20", "20+")
# bucketing data points into bins
TalentData$TotalWorkingYears.Bin <- cut(TalentData$TotalWorkingYears, breaks, include.lowest = T, right=FALSE, labels=labels)


# Monthly-Rate Bins#
# set up boundaries for intervals/bins
breaks <- c(0,5000,10000,15000,20000,25000,Inf)
# specify interval/bin labels
labels <- c("0-5000", "5001-10000","10001-15000","15001-20000","20001-25000","25000+")
# bucketing data points into bins
TalentData$MonthlyRate.Bin <- cut(TalentData$MonthlyRate, breaks, include.lowest = T, right=FALSE, labels=labels)
ac <- table(TalentData$MonthlyRate.Bin)
```

***

#### **Baseline Analysis:**  
The main focus of the EDA was employee attrition rates and what factors may be contributing factors.  

**Baseline Attrition Rate:**
Little Write up here on how we compute it....

```{r AttritionRate}
EmployeeCount.Total = sum(TalentData$EmployeeCount)
AttritionCount.Total = sum(TalentData$AttritionCount)

AttritionRate.Percent = round(AttritionCount.Total / EmployeeCount.Total, 2) * 100 
```




```{r}

## Here's the magic function which computes the Attrition Rates, and Counts
fun.AttritionRate <- function(input.table, group.by.column) {
  input.table %>%
    dplyr::group_by(CategoryVar = input.table[,group.by.column]) %>%
    dplyr::summarise(
      AttritionRate = sum(AttritionCount)/sum(EmployeeCount), 
      AttritionCount = sum(AttritionCount),
      EmployeeCount = sum(EmployeeCount)) %>%
    dplyr::arrange(CategoryVar)
}

## Choose the columns which make sense
ColumnsToAnalyze <- c(
  which( colnames(TalentData)=="NumberOfCompaniesWorked" ),
  which( colnames(TalentData)=="BusinessTravel" ),
  which( colnames(TalentData)=="Department" ),
  which( colnames(TalentData)=="EducationField" ),
  which( colnames(TalentData)=="Gender" ),
  which( colnames(TalentData)=="MaritalStatus" ),
  which( colnames(TalentData)=="JobRole" ),
  which( colnames(TalentData)=="OverTime" ),
  grep("*Label", names(TalentData)), 
  grep("*Bin", names(TalentData))
)


#TalentData <- as_tibble(TalentData)

## Initilize a list to hold all the attrition rate details
Attrition.List = list()

## Loop through all of the columns in TalentData and get the attrition rate details
for(i in ColumnsToAnalyze){
  #Get the name of the current column 
  SummaryColumn <- names(TalentData)[i]
  #Get the attrition rate details into a DF
  SummaryResults <- fun.AttritionRate(TalentData, i)
  #Concatenate the name of the current column to the details
  SummaryResults$ColumnName <- SummaryColumn
  #Store the details into a list with all other column details
  Attrition.List[[i]] <- SummaryResults
}  

## Create a single data frame out of all of the detail results
attrition_data = do.call(rbind, Attrition.List)

```
***

#### Overview of Attrition Rates by each Variable
The following visualizations show the attrition rate of the main variables of the dataset. The height of the bars indicates the attrition percetange for that data point. The width of the bars indicates the overall number of employees who are in those categories.

```{r, fig.width = 8, fig.height= 25}
## Plot
ggplot(data = attrition_data, mapping = aes(x = CategoryVar, y = AttritionRate, width=AttritionCount/100,  fill = AttritionRate)) + 
  geom_bar(stat="identity", position="identity") + 
  scale_colour_gradient2(low = "red", mid = "white",  high = "blue", midpoint = 0, space = "Lab",  na.value = "grey50", guide = "colourbar") +
#  scale_fill_gradient(low='darkgreen', mid='snow3', high='red', space='Lab') +
  geom_hline(yintercept = .16, linetype="dashed", size = .5)+
  geom_text(aes(0,.16,label = "16%", vjust = -.2, hjust = -.1), size=2.5) + 
  geom_text(aes(label=AttritionCount), vjust=1.6, color="white", size=3.5) +
  ylab("Percentage of Attrition") +
  facet_wrap(~ColumnName, scales = 'free_x', ncol = 3) +
  theme(axis.text.x=element_text(angle=90,vjust=1,hjust=1), legend.position="top", plot.title = element_text(hjust = 0.5)) +
  ylab("Percentage of Attrition") +
  ggtitle("Attrition Rates by Variable\n")

```

**Analysis of Age**
Little Write up here on how we compute it....

``` {r}

age.data <- attrition_data[which(attrition_data$ColumnName == "Age.Bin"),]

ggplot(data = age.data, mapping = aes(x = CategoryVar, y = AttritionRate, width=AttritionCount/100,  fill = AttritionRate)) + 
  geom_bar(stat="identity", position="identity") + 
  geom_hline(yintercept = .16, linetype="dashed", size = .5)+
  geom_text(aes(0,.16,label = "16%", vjust = -.2, hjust = -.1), size=3.5) + 
  geom_text(aes(label=paste("n=",AttritionCount)), vjust=-.2, color="navyblue", size=4) +
  theme(axis.text.x=element_text(angle=90,vjust=1,hjust=1), plot.title = element_text(hjust = 0.5)) +
  ylab("Percentage of Attrition") +
  xlab("Age (Binned)") +
  ggtitle("Attrition Rates by Age \n(n/width = sample size)")

```

**Analysis of Job Role**
Little Write up here on how we compute it....

``` {r}

jobrole <- attrition_data[which(attrition_data$ColumnName == "JobRole"),]

ggplot(data = jobrole, mapping = aes(x = reorder(CategoryVar, AttritionRate), y = AttritionRate, width=AttritionCount/100,  fill = AttritionRate)) + 
  geom_bar(stat="identity", position="identity") + 
  geom_hline(yintercept = .16, linetype="dashed", size = .5)+
  geom_text(aes(0,.16,label = "16%", vjust = -.2, hjust = -.1), size=3.5) + 
  geom_text(aes(label=paste("n=",AttritionCount)), vjust=-.2, color="navyblue", size=4) +
  theme(axis.text.x=element_text(angle=90,vjust=1,hjust=1), plot.title = element_text(hjust = 0.5)) +
  ylab("Percentage of Attrition") +
  xlab("Job Role") +
  ggtitle("Attrition Rates by Job Role \n(n/width = sample size)")

```

**Analysis of Job Involvement**
Little Write up here on how we compute it....

``` {r}

job.involvement <- attrition_data[which(attrition_data$ColumnName == "JobInvolvement.Label"),]

job.involvement <- cbind(job.involvement, sortorder = c(3,1,2,4))

ggplot(data = job.involvement, mapping = aes(x = reorder(CategoryVar, sortorder), y = AttritionRate, width=AttritionCount/100,  fill = AttritionRate)) + 
  geom_bar(stat="identity", position="identity") + 
  geom_hline(yintercept = .16, linetype="dashed", size = .5)+
  geom_text(aes(0,.16,label = "16%", vjust = -.2, hjust = -.1), size=3.5) + 
  geom_text(aes(label=paste("n=",AttritionCount)), vjust=-.2, color="navyblue", size=4) +
  theme(axis.text.x=element_text(angle=90,vjust=1,hjust=1), plot.title = element_text(hjust = 0.5)) +
  ylab("Percentage of Attrition") +
  xlab("Job Involvement") +
  ggtitle("Attrition Rates by Job Involvement \n(n/width = sample size)")

```

***
#### Interesting Trends

**Job Involvement vs Satisfaction**
How do these categorical factors relate to each other?

```{r}
Job.Sat.vs.Inv <- TalentData[,c("JobInvolvement", "JobSatisfaction", "JobInvolvement.Label", "JobSatisfaction.Label", "EmployeeCount")]

library(scales)

ggplot(Job.Sat.vs.Inv,aes(x = reorder(JobInvolvement.Label, JobInvolvement), y = EmployeeCount,fill = reorder(JobSatisfaction.Label, -JobSatisfaction))) + 
  geom_bar(position = "fill",stat = "identity") +
  scale_fill_manual(values=c("#045a8d", "#2b8cbe", "#74a9cf", "#bdc9e1")) +
  xlab("Job Involvement") +
  ylab("100 Percent Stacked") +
  ggtitle("Job Involvement vs Satisfaction\n100 Percent Stacked") +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_y_continuous(labels = percent_format())
```

**Job Role ** - Average Age per Role
How do these categorical factors relate to each other?

```{r}

JobRole.Avg.Age <-
  TalentData[,c("JobRole", "Age", "EmployeeCount")] %>%
  dplyr::group_by(JobRole) %>%
  dplyr::summarise(EmployeeCount = sum(EmployeeCount), AverageAge = mean(Age)) 


ggplot(data=JobRole.Avg.Age, aes(x=EmployeeCount, y=AverageAge, color=JobRole, fill=JobRole)) + 
  geom_point(size=5, shape=23)

```


**Job Ratings ** 
How do these categorical factors relate to each other?


```{r}

JobRole.YearsvsIncome <-
  TalentData[,c("JobRole", "YearsInCurrentRole", "MonthlyIncome")] %>%
  dplyr::group_by(JobRole) %>%
  dplyr::summarise(AverageMonthlyIncome = mean(MonthlyIncome), AverageYearsInRole = mean(YearsInCurrentRole)) 


ggplot(data=JobRole.YearsvsIncome, aes(x=AverageYearsInRole, y=AverageMonthlyIncome, color=JobRole, fill=JobRole)) + 
  geom_point(size=5, shape=23)

```


**Job Ratings*** 
How does the attrition patterns differ by Age, Monthly Income broken down by Job Role?
```{r, fig.width = 8, fig.height= 10}

Employee.Performance <- TalentData[,c("EmployeeNumber", "Age", "MonthlyIncome", "JobRole", "Attrition")] 

Employee.Performance$Age <- jitter(Employee.Performance$Age, factor = 3, amount = NULL)


ggplot(data=Employee.Performance, aes(x=Age, y=MonthlyIncome, shape = Attrition, color=Attrition, fill=Attrition)) + 
  geom_point(size=2) + facet_wrap(~JobRole)

```






***
#### Conclusion


